Black-swan events in animal populations

### 2016-11-11 text/ms_pnas.tex
Copy editing

\begin{abstract}
Black swans are improbable events that nonetheless
occur---often with profound consequences.
% While physical extremes, such as
% monsoons and heat waves, are widely studied and increasing in magnitude and frequency,
Such events drive important transitions in social systems (e.g.\ banking
collapses) and physical systems (e.g.\ earthquakes)
yet it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes.
Here we estimate the prevalence and
direction of black-swan events (heavy-tailed process noise)
in \NPops\ animal populations
after accounting for population dynamics (productivity,
density dependence, and typical stochasticity).
We find strong evidence for
black-swan events in \(\sim\)\overallBasePerc \% of populations.
These events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
and are not explained by any life-history covariates, but
tend to be driven by external perturbations such as climate, severe winters,
predators, parasites, or the combined effects of multiple factors.
Black-swan events manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and ignoring heavy-tailed process noise leads to
% a \crashUnderMedian-fold (50\% interval \crashUnderRange)
an underestimate in the magnitude of population crashes.
We suggest modelers consider heavy-tailed downward-skewed
probability distributions,
such as the skewed Student-$t$ used here,
when making forecasts of population abundance.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.
\end{abstract}

@@

Rare catastrophes can have a profound influence on population
persistence \cite{mangel1994}. In recent decades, ecology has moved toward
focusing on aspects of variance in addition to mean
responses \cite{thompson2013}. Our results suggest that an added focus on
ecological extremes represents the next frontier, particularly in the face of
increased climate extremes \cite{meehl2004, ipcc2012, thompson2013}. Financial
analysts are concerned with the shape of downward tails in financial returns
because these directly impact estimates of risk---the probability of a
market crash occurring.
Similarly, ecologists should focus more
on estimating and predicting downward tails of population abundance, since
these increase true extinction risk.

>>>>

\begin{abstract}
Black swans are improbable events that nonetheless
occur---often with profound consequences.
% While physical extremes, such as
% monsoons and heat waves, are widely studied and increasing in magnitude and frequency,
Such events drive important transitions in social systems (e.g.\ banking
collapses) and physical systems (e.g.\ earthquakes)
yet it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes.
Here we estimate the prevalence and
direction of black-swan events (heavy-tailed process noise)
in \NPops\ animal populations
after accounting for population dynamics (productivity,
density dependence, and typical stochasticity).
We find strong evidence for
black-swan events in \(\sim\)\overallBasePerc \% of populations.
These events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
and are not explained by any life-history covariates, but
tend to be driven by external perturbations such as climate, severe winters,
predators, parasites, or the combined effect of multiple factors.
Black-swan events manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and ignoring heavy-tailed process noise leads to
% a \crashUnderMedian-fold (50\% interval \crashUnderRange)
an underestimate in the magnitude of population crashes.
We suggest modelers consider heavy-tailed downward-skewed
probability distributions,
such as the skewed Student-$t$ used here,
when making forecasts of population abundance.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.
\end{abstract}

@@

Rare catastrophes can have a profound influence on population
persistence \cite{mangel1994}. In recent decades, ecology has moved toward
focusing on aspects of variance in addition to mean
responses \cite{thompson2013}. Our results suggest that an added focus on
ecological extremes represents the next frontier, particularly in the face of
increased climate extremes \cite{meehl2004, ipcc2012, thompson2013}. Financial
analysts are concerned with the shape of downward tails in financial returns
because these directly impact estimates of risk---the probability of a
market crash occurring.
Similarly, ecologists should focus more
on estimating and predicting downward tails of population abundance, since
these may increase true extinction risk.
